Scalar Items in Embedded Position: {A}n Experimental Revisit

### 2012-08-07 text/paper/paper1.tex
discussion and conventionalism removed

We consider three main theoretical positions which make different
predictions about the readings of \as- and \es-sentences
\citep[c.f.][for
overview]{Horn2006:The-Border-Wars,Geurts2010:Quantity-Implic,Sauerland2012:The-Computation}. We
will refer to these here as \mymark{traditionalism},
\mymark{conventionalism} and \mymark{grammaticalism} and treat each
one in turn. Since there is some leeway in assessing the predictions
for each of these positions (depending on which of several reasonable
additional assumptions we should adopt), we will distinguish different
varieties of each. For convenience, the predictions of each (variety
of each) position are also summarized in Table~\ref{tab:predictions}
at the end of this section.\dn{micha says: I'll leave all the
  varieties in here until we know which ones we need and which ones we
  don't need}

@@

\subsection{Conventionalism}
\label{sec:conventionalism}

Lexical conventionalism is a position that has been defended for
instance by \citet{LevinsonPresumptiveMeanings2000} and
\citet{Chierchia:2004_ScalarImplicatures}.\fn{See also
  \citet{Sauerland2012:The-Computation} who argues quite favorably for
  a pragmatically informed conventionalism, although eventually
  dismissing it in favor of grammaticalism.} The general idea is that
if the upper-bounding inference from \emph{some} to \emph{some but not
  all} occurs frequently, then it would be inefficient to execute the
Gricean reasoning that traditionalists propose over and over
again. Conventionalism therefore assumes that scalar inferences have
become routinized, hard-wired, so much so that there are two lexical
entries for a scalar item like \emph{some}, one with meaning
\emph{some and maybe all} and one with the meaning \emph{some but not
  all}. \citet{LevinsonPresumptiveMeanings2000} proposed that the
latter is the preferred default meaning, a position which we will
refer to as \mymark{defaultism}.\fn{Levinson's defaultism has strong
  empirical evidence against it, because several processing studies
  show that the computation of scalar inferences appears to be time-
  and effort-consuming, when they \emph{do} arise, not when they
  \emph{don't}, as defaultism would have it
  \citep[c.f.][]{BrehenyKatsos2006:Are-Generalised,BrehenyKatsos2008:Experimental-In,NeysDe-NeysSchaeken2007:When-People-Are}. But
  there is also competing evidence that scalar inferences are
  immediate, low-level processes
  \citep{GrodnerKlein2010:Some-and-Possib}.}\dn{Petra says: empirical
  stuff in footnote irrelevant; I agree, let's move it to discussion
  section} For completeness sake, we will call the opposite view that
the literal meanings are \emph{ceteris paribus} preferred
\mymark{literalism}. We are not aware that anyone has defended this
option, but it pays to include this option, because, as we will see
later on, this possible position is the best direct explanation of our
experimental data.

Both varieties of conventionalism make interesting predictions about
available readings for \as- and \es-sentences. Unaided,
conventionalism does not predict that global readings are
available. The lexically stored upper-bound for scalar \emph{some}
only allows to derive literal and local readings. Moreover, defaultism
predicts that local readings are preferred over literal ones. Contrary
to that, our dummy position of literalism predicts that literal
readings are preferred over local ones.

@@

\begin{table}[t]
  \centering
  \begin{tabular}{lcc}
    & \multicolumn{2}{c}{preference \& availability}
    \\ \cmidrule(r){2-3}
    theoretical position
    & \as
    & \es
    \\ \midrule
    traditionalism
    \\
    \ \ weak restricted 
    & \lit > \glb 
    & \lit > \glb
    \\
    \ \ weak unrestricted
    & \lit > \glb > \loc 
    & \lit > \glb
    \\
    \ \ strong restricted
    & \glb > \lit 
    & \glb > \lit
    \\
    \ \ strong unrestricted
    & \glb > \loc > \lit 
    & \glb > \lit
    \\
    conventionalism
    \\
    \ \ defaultism
    & \loc >  \lit 
    & \loc >  \lit
    \\
    \ \ literalism
    & \lit >  \loc 
    & \lit >  \loc
    \\
    grammaticalism
    \\
    \ \ weak
    & \loc > \glb > \lit 
    & \glb > \lit, \loc
    \\
    \ \ strong
    & \glb, \loc > \lit 
    & \glb, \loc , \lit
    \\
  \end{tabular}
  \caption{Predictions of the relevant theoretical positions. Listed
    are the predicted preference relations. If a reading is not
    listed, it is predicted to be unavailable.}
  \label{tab:predictions}
\end{table}

@@

There is a variation of an existing theory of scalar implicature that
does explain the preference ordering (\ref{bsp:observed-preferences})
in a completely straightforward manner. The theory of interest is
\emph{conventionalism} (also sometimes called \emph{lexicalism})
\citep[e.g.][]{LevinsonPresumptiveMeanings2000,Chierchia:2004_ScalarImplicatures}.
The general idea is that if the upper-bounding inference from
\emph{some} to \emph{some but not all} occurs frequently, then it
would be inefficient to execute the Gricean reasoning that
traditionalists propose over and over again. Conventionalism therefore
assumes that scalar inferences have become routinized, hard-wired, so
much so that there are two lexical entries for a scalar item like
\emph{some}, one with logical meaning \emph{some and maybe all} and
one with the upper-bounded meaning \emph{some but not
  all}. \citet{LevinsonPresumptiveMeanings2000} proposed that the
latter is the preferred default meaning and has therefore ofttimes be
addressed as \mymark{defaultism}. Unfortunately, defaultism has strong
empirical evidence against it, because several processing studies show
that the computation of scalar inferences appears to be time- and
effort-consuming, when they \emph{do} arise, not when they
\emph{don't}, as defaultism would have it
\citep[c.f.][]{BrehenyKatsos2006:Are-Generalised,BrehenyKatsos2008:Experimental-In,NeysDe-NeysSchaeken2007:When-People-Are}. But
there is also competing evidence that scalar inferences are available
in immediate, low-level processes, such as the resolution of
referential expressions in a visual-world paradigm
\citep{GrodnerKlein2010:Some-and-Possib}. For our purposes, it is
interesting to note that the preference relation in
(\ref{bsp:observed-preferences}) is predicted by the dual of
defaultism, call it \emph{literalism}, which posits a lexical
ambiguity but predicts that the logical meanings are \emph{ceteris
  paribus} preferred over the upper-bounded meanings.

Unfortunately, there are problems with conventionalism that seem to
require that it be backed up by traditionalism
\citep[e.g.][]{Sauerland2012:The-Computation}.\fn{\citet{Sauerland2012:The-Computation}
  eventually dismisses even conventionalism backed up by
  traditionalism in favor of grammaticalism based on cases of
  so-called \emph{intermediate implicatures}, where what seems to be
  necessary is the insertion of an exhaustivity operator in
  intermediate scope position \citep[see][for
  details]{Sauerland2012:The-Computation}. Appreciation of such
  intermediate implicatures requires more argumentation, involving a
  discussion of \emph{Hurford's constraint}
  \citep[c.f.][]{FoxSpector:Economy-and-Emb,ChierchiaFox2009:Hurfords-Constr,ChierchiaFox2008:The-Grammatical}.}
For instance, conventionalism cannot account for \emph{indirect
  implicatures}, such as the inference from
(\ref{bsp:indirect-SI-Target}) to (\ref{bsp:indirect-SI-Inference})
which cannot be explained by positing a lexical ambiguity for
\emph{all}.

@@

\noindent In the context of the present study, this would then be good
news for traditionalism, because a combined
traditionalism+conventionalism would predict the observed data
well. The combined position would even be able to explain something
that grammaticalism cannot make sense of as easily, namely the
observation that there were rather consistent localists among the
participants of our study. The combined tradionalism+conventionalism
could explain this by the assumption that some (lazy?, smart?)
speakers have made a conventionalized upper-bounding meaning entry for
scalar \emph{some} while others keep computing things on the
fly. It is not clear how grammaticalism can account for the
intra-subjective consistency.

@@

\begin{itemize}
\item different kinds of ``multiple readings'': syntactic ambiguity
  vs.~pragmatic xxx
\item typicality after all
\end{itemize}

>>>>

We consider two main theoretical positions which make different
predictions about the readings of \as- and \es-sentences
\citep[c.f.][for
overview]{Horn2006:The-Border-Wars,Geurts2010:Quantity-Implic,Sauerland2012:The-Computation}. We
will refer to these here as \mymark{traditionalism} and
\mymark{grammaticalism} and treat each one in turn. Since there is
some leeway in assessing the predictions for each (depending on which
of several reasonable additional assumptions we should adopt), we will
distinguish different varieties of both. For convenience, the
predictions of each (variety of each) position are also summarized in
Table~\ref{tab:predictions} at the end of this section.

@@

@@

\begin{table}[t]
  \centering
  \begin{tabular}{lcc}
    & \multicolumn{2}{c}{preference \& availability}
    \\ \cmidrule(r){2-3}
    theoretical position
    & \as
    & \es
    \\ \midrule
    traditionalism
    \\
    \ \ \ \ weak restricted 
    & \lit > \glb 
    & \lit > \glb
    \\
    \ \ \ \ weak unrestricted
    & \lit > \glb > \loc 
    & \lit > \glb
    \\
    \ \ \ \ strong restricted
    & \glb > \lit 
    & \glb > \lit
    \\
    \ \ \ \ strong unrestricted
    & \glb > \loc > \lit 
    & \glb > \lit
    \\
    % conventionalism
    % \\
    % \ \ defaultism
    % & \loc >  \lit 
    % & \loc >  \lit
    % \\
    % \ \ literalism
    % & \lit >  \loc 
    % & \lit >  \loc
    % \\
    grammaticalism
    \\
    \ \ \ \ weak
    & \loc > \glb > \lit 
    & \glb > \lit, \loc
    \\
    \ \ \ \ strong
    & \glb, \loc > \lit 
    & \glb, \loc , \lit
    \\
  \end{tabular}
  \caption{Predictions of the relevant theoretical positions. Listed
    are the predicted preference relations. If a reading is not
    listed, it is predicted to be unavailable.}
  \label{tab:predictions}
\end{table}

@@

There is a variation of an existing theory of scalar implicature that
does explain the preference ordering (\ref{bsp:observed-preferences})
in a completely straightforward manner. The theory of interest is
\emph{conventionalism} (also sometimes called \emph{lexicalism})
\citep[e.g.][]{LevinsonPresumptiveMeanings2000,Chierchia:2004_ScalarImplicatures}.
The general idea is that if the upper-bounding inference from
\emph{some} to \emph{some but not all} occurs frequently, then it
would be inefficient to execute over and over again the Gricean
reasoning that traditionalists propose. Conventionalism therefore
assumes that scalar inferences have become routinized, hard-wired, so
much so that there are two lexical entries for a scalar item like
\emph{some}, one with logical meaning \emph{some and maybe all} and
one with the upper-bounded meaning \emph{some but not
  all}. \citet{LevinsonPresumptiveMeanings2000} proposed that the
latter is the preferred default meaning, a position which has
therefore ofttimes be addressed as \mymark{defaultism}. Unfortunately,
defaultism has strong empirical evidence against it, because several
processing studies show that the computation of scalar inferences
appears to be time- and effort-consuming, when they \emph{do} arise,
not when they \emph{don't}, as defaultism would have it
\citep[c.f.][]{BrehenyKatsos2006:Are-Generalised,BrehenyKatsos2008:Experimental-In,NeysDe-NeysSchaeken2007:When-People-Are}. But
there is also competing evidence that scalar inferences are available
in immediate, low-level processes, such as the resolution of
referential expressions in a visual-world paradigm
\citep{GrodnerKlein2010:Some-and-Possib}. For our purposes, it is
interesting to note that the preference relation in
(\ref{bsp:observed-preferences}) is predicted by the dual of
defaultism, call it \emph{literalism}, which posits a lexical
ambiguity but predicts that the logical meanings are \emph{ceteris
  paribus} preferred over the upper-bounded meanings.

Unfortunately, there are problems with conventionalism, no matter
which lexical reading is assumed to be preferred, that seem to require
that it be backed up by traditionalism
\citep[e.g.][]{Sauerland2012:The-Computation}.\fn{\citet{Sauerland2012:The-Computation}
  eventually dismisses even conventionalism backed up by
  traditionalism in favor of grammaticalism based on cases of
  so-called \emph{intermediate implicatures}, where what seems to be
  necessary is the insertion of an exhaustivity operator in
  intermediate scope position \citep[see][for
  details]{Sauerland2012:The-Computation}. Appreciation of such
  intermediate implicatures requires more argumentation, involving a
  discussion of \emph{Hurford's constraint}
  \citep[c.f.][]{FoxSpector:Economy-and-Emb,ChierchiaFox2009:Hurfords-Constr,ChierchiaFox2008:The-Grammatical}.}
For instance, conventionalism cannot account for \emph{indirect
  implicatures}, such as the inference from
(\ref{bsp:indirect-SI-Target}) to (\ref{bsp:indirect-SI-Inference})
which cannot be explained by positing a lexical ambiguity for
\emph{all}.

@@

\noindent In the context of the present study, this would then be good
news for traditionalism, because a combined
traditionalism+conventionalism would predict the observed data
well. The combined position would even be able to explain something
that grammaticalism cannot make sense of as easily, namely the
observation that there were rather consistent localists among the
participants of our study. The combined tradionalism+conventionalism
could explain this by the assumption that some (lazy?, smart?)
speakers have made a conventionalized upper-bounding meaning entry for
scalar \emph{some} while others keep computing things on the fly. It
is not clear how grammaticalism can account for the intra-subjective
consistency. Still, this is not an argument \emph{for} traditionalism,
because the point remains that traditionalism as such fails to predict
the observed data.

@@

\begin{itemize}
\item pictures overemphasize contrast some/all
\item subjects tricked into local responses because they are at the end
  \begin{itemize}
  \item would only show availability after all
  \end{itemize}
\item typicality after all
\end{itemize}
